Disclosing funding sources for open access publication fees: the Open APC

### 2016-06-04 article.Rmd
minor fixes

One reason why the study of how much was spent is in most cases difficult is that total publication fee spending is fragmented across the budgets of grant agencies, research institutions, and libraries, or, if support is limited, are taken from personal budgets. Asking 9,645 authors from various disciplines how they financed publication fees, a comprehensive survey in 2010 revealed that the majority of the respondents had access to research funding or institutional support to cover these charges. By contrast, 12 % paid publication fees individually [@soap]. These findings are consistent with that of other studies, adding that sources of funding mostly exists in higher income countries, mainly to support research articles in the bio- and physical sciences [@Solomon_2011]. Personal budgets, on the other hand, are likely used to cover low price publication fees [@Solomon_2011; @Bj_rk_2015].

Another key problem in this regard is that funding for open access journals using publication fees lacks transparency because the parties involved - authors, universities, funders, publishers - neither release information on who pays for what nor the costs of publishing [@Bj_rk_learned_2014], a situation similar to the lack of transparency regarding journal subscriptions [@Lawson_Meghreblian_2015]. Empirical studies examining publication fees so far gathered price information from author surveys [@soap], or journal websites. Using the latter method, two studies investigating journals across multiple disciplines observed similar averages that ranged between 904 $ [@Solomon_2012] and 923 $ [@Walters_2011], as well as considerable price variances across journals and publishers. Accordingly, @Solomon_2012 suggested to cluster fully open access journal using publication fees into several groups. In descending order, these are high-impact journals, followed by biomedicine journals from commercial publishers, large multi-disciplinary journals, and mid-price ranged journals from commercial publishers covering a large spectrum of disciplines. Lower priced journals were published by academic societies and by publishers from low-income countries.

@@

Hybrid journals substantially add to this complexity of open access funding [@Kingsley_2014]. These journals, allowing articles to be published immediately as open access after a charge was paid, rely both on subscriptions and publication fees as revenue sources. Although the uptake of open access through hybrid journals was described as lower and more expensive compared to that of fully open access journals [@Solomon_2012; @Bj_rk_learned_2014; @Pinfield_2015], this model has gained attention through recent science policies, notably because of open access policies from the UK [@Pinfield_Review_2015].

To address these problems of fragmented spending on publication fees and of in-transparency about what was paid, some European research funders and research performing institutions have recently begun to disclose their payment for publication as open data. To our knowledge, the first research funders providing such data were the Wellcome Trust [@wellcome_apc_12] and the Austrian Science Fund FWF [@fwf_apc_13]. The not-for-profit company Jisc followed by collecting data from higher-education institutions in the UK [@Lawson_data_2015]. Disclosed as publicly available spreadsheets, these data-sets self-report what was being spent along with bibliographic information, including title, journal and publisher, and a persistent identifier to the publisher's version. Curatorial efforts focused on the disambiguation of publisher and journal titles as well as on detecting duplicates [@Neylon_2014; @woodward_2014]. Parts of Jisc's cost data was examined by @Pinfield_2015. Although the average spending on publication fees remained stable across the universities, they found large price variances, as well as a varying number of articles UK universities supported between 2007 - 2014, confirming the earlier studies based on price information from journal websites.

@@

This paper focuses on how much German universities and research organisation spent on open access publication fees. In Germany, the Deutsche Forschungsgemeinschaft (DFG), the largest German research funder, has strongly influenced how universities manage institutional support for publication fees.[^3] Before the DFG started to pay for centrally funded publication fees on a pro rata basis through its "Open-Access Publishing" program in 2011, and similar to the situation described in Canada [@Hampson_2014] or the UK [@PINFIELD_2012], only few central funds existed [@eppelin_2012]. Because the DFG has enforced a set of criteria grantees have to comply with, similar implementations for supporting open access publishing across German universities exists [@Fournier_2013]. These criteria exclude sponsoring of articles in hybrid journals, and the funding of articles whose publication fee exceeds 2,000 € (excluding value added tax)[^3]. Grantees agree not only to reimburse the bills on behalf of the researchers they support, but also to look for ways to improve the handling of those financial transactions. They include central invoicing schemes and related agreements between university libraries and publishers, leading to discounts for the authors [@Fournier_2013].

Research institutes organized in the Fraunhofer-Gesellschaft, Helmholtz-Gemeinschaft, Leibniz-Gemeinschaft, and Max-Planck-Gesellschaft are not eligible for this DFG funding program, contributing to the diversity of schemes in Germany. But in response, some organizations have adopted similar processes to support authors. The Max-Planck-Gesellschaft operates their long-lasting open access activities, including handling spending and publisher agreements centrally, through the Max Planck Digital Library [@Schimmer_2013; @Sikora_2015], while the Leibniz-Gemeinschaft set up a dedicated open access fund in 2016.

The evolving institutional support structures to cover open access publication fees has led to calls for an unified approach towards supporting open access journal publishing in Germany. The Allianz der Wissenschaftsorganisationen[^10c], a science policy board representing all major research organizations in Germany, marked cost transparency as one way to sustain an "adequate open access publication system" [@allianz]. Reflecting Austrian and UK initiatives to share institutional spending on open access publication fees as open data, as well as professional discussions on open access publishing, Bielefeld University Library began to openly share its payment of publication fees in May 2014. After engaging with the working group "Electronic Publishing" of the Deutsche Initiative für Netzwerkinformation (DINI)[^11] other German institutions joined under the umbrella of the Open APC initiative soon after.

@@

The aim of the study was to examine how much German universities and research organisations spent on open access publication fees until 2015. Using self-reported cost data from the Open APC initiative, the analysis focused on the amount of institutional payments for publication fees, and compared these expenditure data with data from related Austrian and UK initiatives, both in terms of size and proportion of the different journal types fully and hybrid open access journal. We also asked how thoroughly self-reported articles were indexed in Crossref, a DOI minting agency for scholarly literature, and analysed how institutional spending per articles was distributed over publishers and journal titles.

@@

We analysed self-reported cost data released by the Open APC initiative on May 13, 2016, to assess institutional spending on open access publication fees in Germany. In addition to administrative data about the amount paid per article including value added tax, the reporting institution, and the year of invoicing, we used information about whether an article was published in a fully or hybrid open access journal as well as the recorded DOI from the data-set.

We fetched bibliographic metadata for each article from Crossref on May 19, 2016, on the basis of the reported DOIs. Although the Open APC initiative gathered metadata representing publishers and journals from Crossref as well, this information was retrieved at the time when the participating institutions submitted the cost information. The Open APC initiative tracked the date of data submission and the contributors for every data submission with Git, a version control system, increasingly used for enabling reproducible research [@Ram_2013], and made this information available via GitHub to be transparent over time. However, Crossref regularly updated metadata to represent ongoing mergers of publishing houses or name changes. A prominent example in this regard was the merger of the two large publishing houses Springer Business + Media and Nature Publishing Group announced on May 6, 2015, that operated as Springer Nature at the time of our study. To reflect these dynamics in academic publishing, we decided to retrieve updated metadata from Crossref for the whole Open APC data-set instead of re-using publisher and journal information contained in the Open APC data-set.

As a client, we used the R package rcrossref [@rcrossref], developed and maintained by the rOpenSci initiative[^6], to access Crossref's REST API. We requested the XML-based format `application/vnd.crossref.unixsd+xml` in which full and abbreviated journal titles as well as the ISSN media types, the International Standard Serial Number used to identify journals, were distinguished. It also contained normalized publisher information, thus avoiding confusion about naming of publishing houses other studies were faced with when working with self-reported data [@woodward_2014]. In cases where no bibliographic information could be obtained, we used the Open APC values. Because Crossref was not the only registration agency for DOIs, but also the agencies DataCite and Medra minted DOIs for scholarly work, we furthermore obtained the DOI agency for each article with the help of the rcrossref client.

Data collection also involved obtaining cost data from related open data initiatives. To compare self-reported spending on open access journal articles by Germany universities and research organisations with that of other initiatives, we reviewed the openly available data-sets from the the Austrian Science Fund FWF [@fwf_apc_14; @fwf_apc_15], Jisc [@jisc_14; @jisc_15] and the Wellcome Trust [@wellcome_apc_13_14; @wellcome_apc_14_15]. For analysis, we obtained the overall publication fee spending to support fully and hybrid open access journal articles. In the case of FWF, we gathered the cost information from the accompanying spending reports. We used the spreadsheet data to calculate Wellcome Trust's and Jisc's spending, and converted the prices from GBP to € in accordance with the average € foreign exchange reference rates provided by the European Central Bank. Our comparison between the open data initiatives focussed on the last two years 2014 and 2015. Because Wellcome Trust's spending was reported for the periods 2013 - 2014 and 2014 - 2015, we referred to the average exchange rates of the full two-year period as we could not determine the actual invoicing dates from the data. We excluded articles from the analysis for which either cost information or journal type was missing. In the case of Jisc's 2014 data [@jisc_14], we could not analyze 4.861.772 € spent on 2,812 publications because no publication type was given in the data-set.

The data collection methods of the Open APC initiative and those of the other initiatives differ in some aspects. For instance, whereas the DOI was a mandatory element in the Open APC data template that the participating institutions were required to report, in the Wellcome Trust data a publication identifier was added as part of the automated compliance checks. Our first screening of the data-sets revealed that some articles lacked a DOI. For this reason, and as our main focus is institutional funding for publication fees in Germany, we decided to only compare German APC spending with that reported by other initiatives. We did not compare the overall size, the distribution over publishers and journal titles, or the indexing coverage in Crossref.

@@

After excluding payments being made for non-journal articles and in 2016, we retrieved `r nrow(my_apc)` articles for which the publication fee was centrally paid by `r length(unique(my_apc$institution))` German universities and research institutions until 2015. The number of supported open access journal articles grew over the years (see Figure 1). While one institution disclosed `r filter(my_apc, period == min(my_apc$period)) %>% nrow()` payments made in `r as.character(min(my_apc$period))`, the majority shared their expenditures from 2013 onwards. With `r max(apc_time$articles)` articles, the year `r apc_time[apc_time$articles == max(apc_time$articles),"period"]` was best represented in our data-set. However, `r my_apc %>% filter(period == 2015) %>% group_by(institution) %>% summarise(n()) %>% nrow()` institutions contributed their cost data for 2015 at the time of this analysis, suggesting that there exits a time lag between payments made and reporting these spending to the Open APC initiative.

@@

Among all articles, fees amounted to `r sum(my_apc$euro)` € including VAT, the average payment was `r mean(my_apc$euro)` € and the median value `r median(my_apc$euro)` € .

@@

Figure 3 presents the large price variation among the articles. The disclosed publication fees ranged from `r min(my_apc$euro)` € to `r max(my_apc$euro)` €. However, the average price paid varied somewhat during the period 2011 and 2014 (`r my_apc %>% filter(period == 2011) %>% summarise(mean(euro)) %>% round()` - `r my_apc %>% filter(period == 2014) %>% summarise(mean(euro)) %>% round()` €). We also observe that `r nrow(my_apc[my_apc$euro < 2000,])` (`r (nrow(my_apc[my_apc$euro < 2000,]) / nrow(my_apc)) *100`%) of the publication fees were paid in accordance with the DFG price cap of 2,000 €.

@@

The number of APC payments per institutions varied considerably (see Table 1). With `r filter(my_apc, institution == "MPG") %>% nrow()` reported articles, the Max Planck Society contributed `r (filter(my_apc, institution == "MPG") %>% nrow() / nrow(my_apc)) * 100` % of the overall submissions. In contrast, the universities of technology, TU Clausthal, TU Dortmund TU Ilmenau that recently begun to set up support structures for fee-based open access journal articles, self-reported low numbers of payments.

@@

Table 2 compares Open APC spending data with that of the Austrian FWF, as well as with Jisc's and Wellcome Trust's expenditures. Prices were converted according to the average € exchange rate of the examined periods, and gathered for both fully and hybrid open access journals. The comparison reveals that the Open APC initiative lacked cost information about hybrid journals, whereas the related Austrian and UK open data initiatives could report a large share of spending for these journals between 2014 and 2015. This situation likely reflected the DFG-inspired funding policy of most participating institutions that excluded support for publications in hybrid journals. Over the years 2005 – 2015, `r my_apc %>% group_by(institution) %>% filter(is_hybrid == TRUE) %>% summarise() %>% nrow()` out of `r length(unique(my_apc$institution))` German universities and research institutions reported `r my_apc %>% filter(is_hybrid == TRUE) %>% nrow()` hybrid journal articles to the Open APC initiative, accounting for `r (my_apc %>% filter(is_hybrid == TRUE) %>% nrow() / my_apc %>% nrow()) * 100` % of the overall payments. In terms of the number of supported articles and the amount being spent on publication fees, by contrast, the Open APC data-set provided the most comprehensive price information for fully open access journals compared to what the Austrian and UK initiatives had reported.

@@

Comparison of average prices suggests that publishing in hybrid journal was more expensive than in fully open access journals. Price differentials between these two categories were also reported earlier, concluding that prices for fully open access journals were generally lower [@Solomon_2012; @Pinfield_2015]. In 2014 and 2015, the mean price for fully open access journals calculated from all data-sets was below the DFG price cap of 2,000 €.

@@

To identify the sponsored open access journal articles, as well as to gather bibliographic metadata, DOIs were a mandatory part of the Open APC initiative's data collection activities. The participating institutions reported DOIs for `r my_apc %>% filter(!is.na(doi)) %>% nrow()` out of `r nrow(my_apc)` articles. using these DOIs, we retrieved additional metadata from Crossref for `r nrow(cr)` publications, representing `r nrow(cr) / nrow(my_apc) * 100` % of the total article volume. Articles for which no metadata could be obtained, were either registered with the DOI agency DataCite (10 articles) or Medra (two articles). For eight articles, our parser could not gather the XML resource, although these publications were registered with Crossref. Seven DOIs reported to the Open APC initiative did not resolve.

@@

We used the DOI to automatically fetch publisher and journal names for each article from the Crossref REST API. Table 4 shows the top ten publishers in terms of payments made that represent `r apc_publisher %>% filter(!publisher == "Other") %>% nrow() / my_apc %>% nrow() * 100` % of the spending for publication fees. In total, payments were made to `r my_apc %>% group_by(publisher) %>% summarize(n()) %>% nrow()` publishing houses. In comparison with data from the UK, full open access publishers have a greater share on total spending. @Pinfield_2015, for instance, reported remarkably lower numbers for the open access publishers MPDI AG, Copernicus GmbH, and Hindawi Publishing.

@@

Most of the funding of publication fees in Germany went to the publisher Springer Nature, most llikeky profiting from the merge with the open access publisher BioMed Central in 2008, and in 2015 between the well-established publishers Springer Science + Business Media and Nature Publishing Group. Using the Crossref-Member-ID instead of the publisher name, we were able to differentiate between journals formely published by Springer Science + Business Media and Nature Publishing Group. In terms of articles, the majority of payments made were for publications in journal that were formerly associated with Springer Science + Business Media. Springer Science + Business Media journals accounted for `r nrow(springer)`articles, representing `r nrow(springer) / nrow(springer_nat) * 100` of the total Springer Nature article volume in the spending data, and `r sum(springer$euro) / sum(springer_nat$euro) * 100` % of the amount spent. Median publication fee spending slightly differed between Springer Science + Business Media (`r median(springer$euro)` €) and Nature Publishing (`r median(nature$euro)` €). However, price variation was higher for Nature Publishing journals (SD = `r sd(nature$euro)`€) compared with that of the former Springer Science + Business Media titles (SD = `r sd(nature$euro)` €).

In contrast to Springer Nature, other well-established publishing houses such as Elsevier and Wiley-Blackwell rank lower in our analysis, presumably because they mostly published hybrid journals, a journal type that was not well represented in our data-set at the time of this study. Table 4 also illustrates the variation across and within publishers, which confirms earlier findings [@Solomon_2012; @Pinfield_2015].

@@

Prices also varied within single journals. Based on the number of articles paid for, Table 5 illustrates the top ten out of `r my_apc %>% group_by(journal_full_title) %>% summarize(n()) %>% nrow()` journals. Payments to these ten journals represent `r apc_journal %>% filter(!journal_full_title == "Other") %>% nrow() / my_apc %>% nrow() * 100` % of all payments. In the case of Atmospheric Chemistry and Physics Discussions, the price range can be explained by the fact that this journal charges per page and also takes the submission's file format into consideration.

@@

>>>>

One reason why the study on how much was being spent is in most cases difficult is that total publication fee spending is fragmented across the budgets of grant agencies, research institutions, and libraries, or, if support is limited, are taken from personal budgets. Asking 9,645 authors from various disciplines how they financed publication fees, a comprehensive survey in 2010 revealed that the majority of the respondents had access to research funding or institutional support to cover these charges. By contrast, 12 % paid publication fees individually [@soap]. These findings are consistent with that of other studies, adding that sources of funding mostly exists in higher income countries, mainly to support research articles in the bio- and physical sciences [@Solomon_2011]. Personal budgets, on the other hand, are likely used to cover low price publication fees [@Solomon_2011; @Bj_rk_2015].

Another key problem in this regard is that funding for open access journals using publication fees lacks transparency because the parties involved - authors, universities, funders, publishers - neither release information on who pays for what nor the costs of publishing [@Bj_rk_learned_2014], a situation similar to the lack of transparency regarding journal subscriptions [@Lawson_Meghreblian_2015]. Empirical studies examining publication fees so far obtained price estimates by surveying authors [@soap], or from journal websites. Using the latter method, two studies investigating journals across a broad range of disciplines calculated similar price averages that ranged between 904 $ [@Solomon_2012] and 923 $ [@Walters_2011], as well as considerable price variances across journals and publishers. Accordingly, @Solomon_2012 suggested to cluster fully open access journal using publication fees into several groups. In descending order, these are high-impact journals, followed by biomedicine journals from commercial publishers, large multi-disciplinary journals, and mid-price journals from commercial publishers covering a large spectrum of disciplines. Lower priced journals were published by academic societies and by publishers from low-income countries.

@@

Hybrid journals substantially add to this complexity of open access funding [@Kingsley_2014; @Pinfield_2015]. These journals, allowing articles to be published immediately as open access after a charge was paid, rely both on subscriptions and publication fees as revenue sources. Although the uptake of open access through hybrid journals was described as lower and more expensive compared to that of fully open access journals [@Solomon_2012; @Bj_rk_learned_2014], this model has gained attention through recent science policies, notably because of open access policies from the UK [@Pinfield_Review_2015].

To address these problems of fragmented spending on publication fees and of in-transparency about what was being paid, some European research funders and research performing institutions have recently begun to disclose their payment for publication as open data. To our knowledge, the first research funders providing such data were the Wellcome Trust [@wellcome_apc_12] and the Austrian Science Fund FWF [@fwf_apc_13]. The not-for-profit company Jisc followed by collecting data from higher-education institutions in the UK [@Lawson_data_2015]. Disclosed as publicly available spreadsheets, these data-sets self-report what was being spent along with bibliographic information, including title, journal and publisher, and a persistent identifier to the publisher's version. Curatorial efforts focused on the disambiguation of publisher and journal titles as well as on detecting duplicates and persistent identifiers to the full text [@Neylon_2014; @woodward_2014]. Parts of Jisc's cost data was examined by @Pinfield_2015. Although the average spending on publication fees remained stable across the universities, they found large price variances, as well as a varying number of articles UK universities supported between 2007 - 2014, confirming earlier studies that collected price information from journal websites.

@@

This paper focuses on how much German universities and research organisation spent on open access publication fees. In Germany, the Deutsche Forschungsgemeinschaft (DFG), the largest German research funder, has strongly influenced how universities manage institutional support for publication fees.[^3] Before the DFG started to pay for centrally funded publication fees on a pro rata basis through its "Open-Access Publishing" program in 2011, and similar to the situation described in Canada [@Hampson_2014] or the UK [@PINFIELD_2012], only few central funds existed [@eppelin_2012]. Because the DFG has enforced a set of criteria grantees have to comply with, similar implementations for supporting open access publishing across German universities exists [@Fournier_2013]. These criteria exclude sponsoring of articles in hybrid journals, and the funding of articles whose publication fee exceeds 2,000 € (excluding value added tax)[^3]. Grantees agree not only to reimburse the bills on behalf of the researchers they support, but also to look for ways to improve the handling of those financial transactions. This includes central invoicing schemes, and related agreements between university libraries and publishers, leading to discounts for the authors [@Fournier_2013].

Research institutes organised in the Fraunhofer-Gesellschaft, Helmholtz-Gemeinschaft, Leibniz-Gemeinschaft, and Max-Planck-Gesellschaft are not eligible for this DFG funding program. But in response, some organizations have adopted similar processes to support authors. The Max-Planck-Gesellschaft operates their long-lasting open access activities, including handling spending and publisher agreements centrally, through the Max Planck Digital Library [@Schimmer_2013; @Sikora_2015], while the Leibniz-Gemeinschaft set up a dedicated open access fund in 2016.

The evolving institutional support structures to cover open access publication fees has led to calls for an unified approach towards supporting open access journal publishing in Germany. The Allianz der Wissenschaftsorganisationen[^10c], a science policy board representing all major research organisations in Germany, marked price transparency as one way to sustain an "adequate open access publication system" [@allianz]. Reflecting Austrian and UK initiatives to share institutional spending on open access publication fees as open data, as well as professional discussions on open access publishing, Bielefeld University Library began to openly share its payment of publication fees in May 2014. After engaging with the working group "Electronic Publishing" of the Deutsche Initiative für Netzwerkinformation (DINI)[^11] other German institutions joined under the umbrella of the Open APC initiative soon after.

@@

The aim of the study was to examine how much German universities and research organisations spent on open access publication fees until 2015. Using self-reported cost data from the Open APC initiative, the analysis focused on the amount that was being spent on publication fees, and compared these expenditure with data from related Austrian and UK initiatives, both in terms of size and the proportion of articles being published in fully and hybrid open access journals. We also asked how thoroughly self-reported articles were indexed in Crossref, a DOI minting agency for scholarly literature, and analysed how institutional spending per articles was distributed over publishers and journal titles.

@@

We analysed self-reported cost data released by the Open APC initiative on May 13, 2016[^24] to assess institutional spending on open access publication fees in Germany. In addition to administrative data about the amount paid per article including value added tax, the reporting institution, and the year of invoicing, we used information about whether an article was published in a fully or hybrid open access journal as well as the recorded DOI from the data-set.

We fetched bibliographic metadata for each article from Crossref on May 19, 2016, on the basis of the reported DOIs. Although the Open APC initiative gathered metadata representing publishers and journals from Crossref as well, this information was retrieved at the time when the participating institutions submitted their expenses. The Open APC initiative kept track of the date when these data-sets were submitted with Git, a version control system, increasingly used for enabling reproducible research [@Ram_2013], and made this information available via GitHub to be transparent over time. However, during these data collection activities Crossref had regularly updated metadata to represent ongoing mergers of publishing houses. A prominent example in this regard was the merger of the two large publishing houses Springer Business + Media and Nature Publishing Group announced on May 6, 2015, that operated as Springer Nature at the time of our study. To reflect these dynamics in academic publishing, we decided to retrieve updated metadata from Crossref for the whole Open APC data-set instead of re-using the historical publisher and journal information contained in the Open APC data-set.

As a client, we used the R package rcrossref [@rcrossref], developed and maintained by the rOpenSci initiative[^6], to access Crossref's REST API[^25]. We requested the XML-based format `application/vnd.crossref.unixsd+xml` in which full and abbreviated journal titles as well as the ISSN media types, the International Standard Serial Number used to identify journals, were distinguished. It also contained normalised publisher information, thus avoiding confusion about naming of publishing houses other studies were faced with when working with self-reported data [@woodward_2014]. In cases where no bibliographic information could be obtained, we used the Open APC values. Because Crossref was not the only registration agency for DOIs, but also the agencies DataCite and Medra minted DOIs for scholarly work, we furthermore obtained the DOI agency for each article with the help of the rcrossref client.

Data collection also involved obtaining cost data from related open data initiatives. To compare self-reported spending on open access journal articles by Germany universities and research organisations with that of other initiatives, we consulted the openly available data-sets from the the Austrian Science Fund FWF [@fwf_apc_14; @fwf_apc_15], Jisc [@jisc_14; @jisc_15] and the Wellcome Trust [@wellcome_apc_13_14; @wellcome_apc_14_15]. For analysis, we obtained the overall publication fee spending on both fully and hybrid open access journal articles. In the case of FWF, we gathered the cost information from the accompanying spending reports. We used the spreadsheet data to calculate Wellcome Trust's and Jisc's spending, and converted the prices from GBP to € in accordance with the average € foreign exchange reference rates provided by the European Central Bank. Our comparison between the open data initiatives focussed on the last two years 2014 and 2015. Because Wellcome Trust's spending was reported for the financial periods 2013 - 2014 and 2014 - 2015, we referred to the average exchange rates of the full two-year period as we could not determine the actual invoicing dates from the data. We excluded articles from the analysis for which either cost information or journal type was missing. In the case of Jisc's 2014 data [@jisc_14], for instance, we excluded spending on 2,812 publications that amounted to 4.861.772 € from the analysis because no publication type was given in the data-set.

The data collection methods of the Open APC initiative and those of the other initiatives differed in some aspects. For instance, whereas the DOI was a mandatory element in the Open APC data template that the participating institutions were required to report, in the Wellcome Trust data a publication identifier was added as part of the automated compliance checks. Our first screening of the data-sets revealed that some articles lacked a DOI. For this reason, and as our main focus was institutional funding for publication fees in Germany, we decided to only compare German APC spending with that reported by other initiatives. We did not compare the distribution of spending over publishers and journal titles, or the indexing coverage in Crossref.

@@

After excluding payments being made for non-journal articles, as well as articles invoiced in 2016, we retrieved `r nrow(my_apc)` open access journal articles `r length(unique(my_apc$institution))` German universities and research institutions financially supported between 2005 and 2015. As illustrated in Figure 1, payments made for open access journal articles grew over the years. While one institution supported `r filter(my_apc, period == min(my_apc$period)) %>% nrow()` articles in `r as.character(min(my_apc$period))`, the majority shared their expenditure from 2013 onwards. With `r max(apc_time$articles)` articles, the year `r apc_time[apc_time$articles == max(apc_time$articles),"period"]` was best represented in our data-set. However, the data-set contained `r my_apc %>% filter(period == 2015) %>% group_by(institution) %>% summarise(n()) %>% nrow()` institutions that contributed their cost data for 2015 at the time of this analysis, suggesting a time lag between payments made and reporting these spending to the Open APC initiative.

@@

Among all articles, fees amounted to `r sum(my_apc$euro)` € including VAT, the average payment was `r mean(my_apc$euro)` € and the median value `r median(my_apc$euro)` € . Figure 2 presents the distribution of institutional spending on publications. We observed that `r nrow(my_apc[my_apc$euro < 2000,])` (`r (nrow(my_apc[my_apc$euro < 2000,]) / nrow(my_apc)) *100`%) of the publication fees were paid in accordance with the DFG price cap of 2,000 €. Most payment on publications ranged between 1,000 - 1,250 €.

@@

Figure 3 presents the large price variations. Publication fees that were paid by the German universities and research organisations ranged between `r min(my_apc$euro)` € and `r max(my_apc$euro)` €  (SD `r sd(my_apc$euro)`). However, the average price paid varied somewhat during the period 2011 and 2015 (`r my_apc %>% filter(period == 2011) %>% summarise(mean(euro)) %>% round()` - `r my_apc %>% filter(period == 2015) %>% summarise(mean(euro)) %>% round()` €). Average publication fee spending per article in Germany was `r mean(my_apc$euro)` (Median = `r median(my_apc$euro)` €, SD = `r sd(my_apc$euro)` €).

@@

The number of APC payments per institutions varied considerably (see Table 1). With `r filter(my_apc, institution == "MPG") %>% nrow()` reported articles, the Max Planck Society contributed `r (filter(my_apc, institution == "MPG") %>% nrow() / nrow(my_apc)) * 100` % of the overall submissions. By contrast, the universities of technology, TU Clausthal, TU Dortmund TU Ilmenau that recently begun to set up support structures for fee-based open access journal articles, self-reported low numbers of payments.

@@

Table 2 compares Open APC spending data with that of the Austrian FWF, as well as with Jisc's and Wellcome Trust's expenditure. Prices were converted according to the average Euro exchange rate of the examined periods, and gathered for both fully and hybrid open access journals. The comparison reveals that the Open APC initiative lacked cost information about hybrid journals, whereas the related Austrian and UK open data initiatives could report a large share of spending on these journals between 2014 and 2015. This situation presumably reflected the importance of the DFG as funding source for publication fees in Germany. The DFG funding policy excluded support for publications in hybrid journals. Over the years 2005 – 2015, `r my_apc %>% group_by(institution) %>% filter(is_hybrid == TRUE) %>% summarise() %>% nrow()` out of `r length(unique(my_apc$institution))` German universities and research institutions reported `r my_apc %>% filter(is_hybrid == TRUE) %>% nrow()` hybrid journal articles to the Open APC initiative, representing `r (my_apc %>% filter(is_hybrid == TRUE) %>% nrow() / my_apc %>% nrow()) * 100` % of the overall supported article volume. In terms of the number of supported articles and the amount being spent on publication fees, by contrast, the Open APC data-set provided the most comprehensive price information for fully open access journals compared to what the Austrian and UK initiatives had reported.

@@

Comparison of average prices suggests that publishing in hybrid journal was more expensive than in fully open access journals. Price differentials between these two categories were also reported earlier, indicating that prices for fully open access journals were typically lower [@Solomon_2012; @Pinfield_2015]. In 2014 and 2015, the mean price for fully open access journals calculated from all data-sets was below the DFG price cap of 2,000 €.

@@

To identify publication fee spending on the article-level, as well as to gather bibliographic metadata, DOIs were a mandatory part of the Open APC initiative's data collection activities. The participating institutions reported DOIs for `r my_apc %>% filter(!is.na(doi)) %>% nrow()` out of `r nrow(my_apc)` articles. Using these DOIs, we retrieved additional metadata from Crossref for `r nrow(cr)` publications, representing `r nrow(cr) / nrow(my_apc) * 100` % of the total article volume. Articles for which no metadata could be obtained, were registered with the DOI agency DataCite (10 articles) or Medra (two articles). For eight articles, our parser could not gather the XML resource, although these publications were registered with Crossref at the time of our study. Seven DOIs reported to the Open APC initiative did not resolve.

@@

We used the DOI to automatically fetch publisher and journal names for each article from the Crossref REST API. Table 4 shows the top ten publishers in terms of payments made that represented `r apc_publisher %>% filter(!publisher == "Other") %>% nrow() / my_apc %>% nrow() * 100` % of the overall spending on publication fees. In total, payments were made to `r my_apc %>% group_by(publisher) %>% summarize(n()) %>% nrow()` publishing houses. In comparison with data from the UK, some open access publishers have a greater share on total spending. @Pinfield_2015, for instance, reported remarkably lower numbers for the open access publishers MPDI AG, Copernicus GmbH, and Hindawi Publishing.

@@

Most of publication fee spending in Germany was on articles published in Springer Nature journals, most likely profiting from the merge with the open access publisher BioMed Central in 2008, and that between the well-established publishers Springer Science + Business Media and Nature Publishing Group in 2015. Using the Crossref-Member-ID instead of the publisher name, we were able to differentiate between journals formely published by Springer Science + Business Media and Nature Publishing Group. In terms of articles, the majority of payments made were for publications in journals formerly associated with Springer Science + Business Media. Springer Science + Business Media journals accounted for `r nrow(springer)` articles, representing `r nrow(springer) / nrow(springer_nat) * 100`% of the overall Springer Nature article volume in the Open APC data, and `r sum(springer$euro) / sum(springer_nat$euro) * 100` % in terms of the amount being spent. Median publication fee spending slightly differed between Springer Science + Business Media (`r median(springer$euro)` €) and Nature Publishing (`r median(nature$euro)` €). However, price variation was higher for Nature Publishing journals (SD = `r sd(nature$euro)`€) compared with that of the former Springer Science + Business Media titles (SD = `r sd(springer$euro)` €).

In contrast to Springer Nature, other well-established publishing houses such as Elsevier and Wiley-Blackwell rank lower in our analysis. Table 2, furthermore, illustrates the price variations across and within publishers, which confirms earlier findings [@Solomon_2012; @Pinfield_2015].

@@

Prices also varied within single journals. Based on the number of articles paid for, Table 3 illustrates the top ten out of `r my_apc %>% group_by(journal_full_title) %>% summarize(n()) %>% nrow()` journals. Payments to these journals represent `r apc_journal %>% filter(!journal_full_title == "Other") %>% nrow() / my_apc %>% nrow() * 100` % of the overall expenditure. In the case of Atmospheric Chemistry and Physics Discussions, the large price range can be explained by the fact that this journal charges per page and also takes the submission's file format into consideration.

@@

[^24]: https://github.com/OpenAPC/openapc-de/tree/v2.4.3

[^25]: <https://github.com/CrossRef/rest-api-doc/blob/master/rest_api.md>
